---
title: "Characterizing Neighborhood Using 311 Calls: \n A Case Study of Kansas City"
output: github_document
---
<center>

Thao Tran

October 2023 

</center>

<!-- **Abstracts** -->

<!-- TBD  -->


### Introduction

Urban environments are characterized by a complex tapestry of diverse neighborhoods, each exhibiting unique socio-economic, cultural, and infrastructural features. Effective urban planning and decision-making demand an intricate understanding of these localized contexts. Traditional demographic estimates, though invaluable, often suffer from temporal lags that limit their real-time applicability. In response to this challenge, emerging research suggests that 311 non-emergency service calls offer an innovative avenue for constructing a dynamic and cost-effective indicators of urban neighborhoods. These service requests, encompassing a wide array of concerns raised by residents, possess the potential to serve as invaluable indicators of neighborhood distress and characteristics. By scrutinizing this wealth of data, urban stakeholders can gain unprecedented insights into local conditions, aiding in the formulation of targeted interventions and policies.

The utilization of 311 Service Requests as a tool for characterizing urban neighborhoods represents a burgeoning area of inquiry in urban studies and planning. Notably, seminal work by [Wang et al (2017)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5645100/) highlights the capacity of 311 data to unveil the intricate fabric of urban systems. Through detailed analysis of New York City, Boston, and Chicago, the study demonstrates how these requests, categorized by type and location, furnish a robust foundation for delineating distinct socio-economic profiles across neighborhoods. This innovative classification, hinged on 311 data, presents an invaluable resource for modeling various socio-economic features, offering a real-time counterpart to conventional, lagged estimates.

Furthermore, the implications of this research extend beyond theoretical classifications. The findings underscore the practical utility of 311-based characterizations in forecasting future trends, particularly in the realm of local real estate prices. This revelation suggests that 311 Service Requests harbor the potential not only to monitor but also to predict the socio-economic performance of urban neighborhoods. Such predictive capabilities empower urban stakeholders with the means to quantitatively evaluate the impacts of their interventions, fostering a more dynamic and responsive approach to urban planning.

In light of these compelling insights, this paper builds upon this foundational work, delving deeper into the potential of 311 data as a dynamic lens through which to understand and address the challenges of urban environments. Through rigorous analysis and modeling, we seek to augment the understanding of how 311 Service Requests in Kansas City Missouri (KCMO) can serve as an invaluable real-time indicator of neighborhood characteristics, complementing conventional demographic estimates and enabling more informed, timely, and effective urban interventions.

```{r setup, include=FALSE}
#load library
library(tidyverse)
options(dplyr.summarise.inform =FALSE)
library(reshape2)
library(openxlsx)
library(car)
library(ggplot2)
library(hrbrthemes)
library(viridis)
#load library for word association graph
library(magrittr)
library(purrr)
library(tibble)
library(tidyr)
library(tidytext)
library(igraph)
library(ggraph)
library(corpus)
library(janitor)
library(knitr)
library(rgdal)
library(sf)
library(scales)
library(fmsb)
library(scales)
library(useful)
#set colors 
colors_hex = c("dodgerblue4","green3","red3","orange2","purple4","yellow", "sienna4","hotpink","paleturquoise3") 

#load data
load("data/kcmo2019_2020_sec.rdata")

#create year+month var
dat$yrmo = dat$CREATEYR*100 +dat$CREATEMO
dat$date = as.Date(as.character(dat$yrmo*100+1),"%Y%m%d")

#subset 311 data from March 2019 to February 2020 to avoid pandemic and remove NA category
dat = dat %>% filter(yrmo %in% 201903:202002 & CATEGORY != "Data Not Available")

#load socio-demo data
sec = read.xlsx("data/SocioEconomic.xlsx") %>%
  mutate(
    Est.Population = Number.of.Households*3, #Average household size in KCMO is 3
    Share.Bachelors.degree.or.higher = 100*`Total.population.age.25+.years.with.a.bachelor's.degree.or.higher`/Est.Population,
    Share.White.alone = 100*White.alone/Est.Population,
    Share.Black.or.African.American.alone= 100*Black.or.African.American.alone/Est.Population,                                  
    Share.Asian.alone = 100*Asian.alone/Est.Population,                                                       
    Share.Hispanic.or.Latino = 100*Hispanic.or.Latino/Est.Population,                                               
    Share.Vacant = 100*Total.Vacant/Number.of.Households,
    Share.Renter.Occupied = 100*Total.Renter.Occupied/Number.of.Households                                             
  ) %>% select(
    NBH_ID,
    Median.income,
    Median.home.value,
    Share.Bachelors.degree.or.higher,
    Share.White.alone:
    Share.Renter.Occupied,
    Number.of.Households
  )

#load shp file
shp <- readOGR(dsn = "data/Kansas City Neighborhood Boundaries/geo_export_81808d3e-44c5-48c1-9a8e-76fec558c110.shp", stringsAsFactors = F)
# convert to sf
shp_sf = st_as_sf(shp)
```

### Data

#### About 311 data

The data utilized in this study originates from the Kansas City Open Data website, spanning from March 1st, 2019 to February 1st, 2020. This temporal scope allows for an examination of non-emergency service requests across four counties within the Kansas City, Missouri area, namely Jackson, Clay, Platte, and Cass. This pre-pandemic timeframe offers a valuable baseline for understanding the dynamics of urban service requests prior to the influence of extraordinary events such as the COVID-19 pandemic. The 311 system serves as a conduit through which local government entities provide non-emergency assistance to residents, visitors, and businesses. It addresses reported disruptions in services, unsafe situations, and disturbances affecting quality of life. Although requests are received through various channels such as text messages, web forms, and dedicated mobile applications, the majority are received via phone calls. These 311 service requests and complaints encompass a wide spectrum of concerns, including, but not limited to, noise disturbances, building heating outages, and sightings of rodents. Consequently, this data proves to be an invaluable resource for comprehending the provision of crucial municipal services and gauging neighborhood conditions.

```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.height=8,fig.width=8, dpi=300}
#add 311 volume to geo data
vol311 = dat %>% group_by(nbhid = NBH_ID) %>% summarise(vol311=n())

#map 311 vol by nbh
shp_sf %>% 
  left_join(vol311, by = "nbhid") %>%
  ggplot() +
  geom_sf(aes(fill = vol311 ), color = "black")+
  labs(fill = "Total 311 Calls Volume", title = "Figure 1: 311 Calls Volume by Neighborhood in Kansas City, MO from March 2019 to february 2020 ")+
  theme_void()
```

The number of total requests during the time frame for the 311 data in KCMO is at 117 thousands, about a fifth of the city's population during the same year. Kansas City's 311 data are categorized in 15 types which are listed below.
 
```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.height=8,fig.width=8, dpi=300}

#summarize 311 volume on NBH level
dat %>%group_by(CATEGORY, NEIGH) %>% 
  summarise(n = n()) %>% 
  ungroup() %>%
  group_by(CATEGORY) %>% 
  summarise(
    Neighborhood_Count = n(),
    Min = min(n),
    Mean = mean(n),
    Median = median(n),
    SD = sd(n),
    Max = max(n),
    Total = sum(n)
  ) %>% kable(digits = 0,format.args = list(big.mark = ","),
              caption = "Table 1: Summary statitics of 311 Calls by 311 Non-Emergency Service Categories")

```

The 311 data contains one record for each call including information such as: type of request, time of request, open/close status of request, and location of request. We aggregate 311 calls and group by type and neighborhood. 

#### About demographic and socioeconomic data

As we attempt to use 311 data to characterize urban neighborhood in complement to existing socio-demographic indicators. For socioeconomic and demographic features, we use data data from U.S. Census 2019 American Community Survey (ACS).
 
```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.height=8,fig.width=8, dpi=300}

# summarise demographic and socioeconomic data
data.frame(t(sec %>% select(-c(NBH_ID)))) %>% 
  add_rownames("Socioeconomic_Demographic") %>% 
  melt() %>% 
  group_by(Socioeconomic_Demographic) %>% 
  summarise(
    Neighborhood_Count = n(),
    Min = min(value),
    Mean = mean(value),
    Median = median(value),
    SD = sd(value),
    Max = max(value),
  ) %>% 
  kable(digits = 2,format.args = list(big.mark = ","),
        caption = "Table 2: Summary statistics of socio-demographic characteristics across neighborhoods in Kansas City, MO")
```

### Classification based on 311 service categories

We define a 311 service composition for a neighborhood using the relative frequencies of 311 calls across different request types. These compositions allow us to gain initial insight to the usage and needs of 311 services across different communities. This approach characterizes the unique residents' needs for 311 service in a given neighborhood, highlighting particular concerns that are most important to the local community. 

For a neighborhood $i$, its composition is represented with $C(i) = (\frac{N_{i,t}}{N_i}, t=1...T)$, where $N_{i,t}$ is the total number of request in category $t$ in neighborhood $i$, $N_i$ is the total number of 311 request in neighborhood $i$ which can be also be written as $N_i = \sum_t N_{i,t}$, and $T$ is total number of service categories.

The 311 services composition $C_i$ describes unique characteristics of neighborhood $i$ and there are likely groups of neighborhoods that share similar composition like how areas share similar socioeconomic characteristics. We apply k-mean clustering to multi-dimensional composition $C_i$ to classify neighborhoods with similar 311 service characteristics into distinct clusters. We get an optimal clustering we run the algorithm 100 times using cumulative square sum of distances from centroids.  

The key step in k-mean clustering is to identify an appropriate number of clusters. We evaluate the clustering model with Hartigan–Wong method as shown below. Given the socioeconomic diversity across neighborhoods in the selected cities, we determine that a minimum of three clusters is an appropriate value.

```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.height=8,fig.width=8, dpi=300}
# prepare data for clustering
vol311_nbh = dat %>% 
  group_by(NBH_ID,NEIGH) %>%
  mutate(N = n()) %>%
  group_by(NBH_ID, NEIGH,CATEGORY) %>%
  summarise(n=n(), N= max(N),c_it = n/N) %>% 
  select(-c(n,N))%>%
  dcast(NBH_ID+NEIGH~CATEGORY)%>% 
  replace(is.na(.), 0)

#choose number of clusters
kBest = FitKMeans((vol311_nbh[,-c(1:2)]), max.clusters=20, nstart=25, seed=20231005)
kBest %>% kable()
PlotHartigan(kBest) + theme_bw() +
  labs(title = "Figure 2: Hartigan's Rule to determine k")
```

In the following figure, we show how the neighborhood are divided into seven clusters base on the 311 service composition. 

```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.height=8,fig.width=8, dpi=300}
#k
k = 7

#k means cluster
set.seed(20231005)
K3N25 = kmeans((vol311_nbh[,-c(1:2)]), centers=k, nstart=25) # see the

#add cluster to nbh ID 
K3N25_label = data.frame(nbhid = vol311_nbh$NBH_ID, Neighborhood = vol311_nbh$NEIGH,Cluster = K3N25$cluster)

#list NBH names by cluster
for (i in 1:k){
  cat("- Cluster", i, "includes:", paste(K3N25_label$Neighborhood[K3N25_label$Cluster==i],","),"\n\n")
}


#visualize cluster on the map
shp_sf %>% 
  left_join(K3N25_label, by = "nbhid") %>%
  ggplot() +
  geom_sf(aes(fill = as.factor(Cluster)), color = "black")+
  labs(fill = "Cluster", title = "Figure 3: K-Means classication of neighborhoods based on 311 call volumes")+
  scale_fill_manual(values = colors_hex)+
  theme_void()

```

To gauge the difference in 311 service request characteristics among clusters, we illustrate the distribution of top service requests categories across the seven clusters. This reveals a distinct variation in the composition.

```{r, echo=FALSE, warning=FALSE,message=FALSE,fig.height=8,fig.width=8, dpi=300}
# plot share of top categories for each cluster
dat %>% 
  left_join(K3N25_label, by = c("NBH_ID" = "nbhid")) %>% 
  group_by(Cluster, CATEGORY) %>%
  summarise(n= n()) %>% 
  group_by(Cluster) %>%
  mutate(
    total = sum(n),
    share = n/total
  ) %>% 
  slice_max(order_by = share, n = 5) %>%
  ggplot(aes(y = CATEGORY,  x = share, fill = CATEGORY)) +
  geom_col(stat = "identity", position = position_dodge(),show.legend = FALSE)+
  facet_wrap(~Cluster, nrow =5, scales = "free_y")+
  labs(y = "", x = "", title = "Figure 4: Top 311 service categories by KCMO neighborhood clusters")+
  scale_x_continuous(labels = percent)+
  scale_fill_manual(values = colors_hex)+
  theme_minimal()

  
  
```


### Socioeconomic features among clusters

Given our understanding of the specific spatial contexts within the analyzed cities, the emerging clusters align with intuitive expectations. However, to quantitatively address how the local 311 service request patterns resembles the socioeconomic attributes of those areas, we now compile and scrutinize the socioeconomic characteristics associated with each of the identified clusters.

It's important to note that, up to this point, the clustering outcomes have been derived solely from the frequency of 311 service requests, without incorporating any socioeconomic data. In the following step, we aggregate information regarding socioeconomic indicators such as income levels, educational attainment, as well as racial composition. We then compare the ranking (ascending) of average levels of each indicator within each of the designated clusters.

```{r, echo=FALSE, warning=FALSE,message=FALSE,fig.height=8,fig.width=8, dpi=300}

#radar chart to show socioeconomic by cluster

K3N25_label %>% 
  left_join(sec,by = c("nbhid"="NBH_ID")) %>% 
  select(-c(nbhid,Neighborhood))%>%
  group_by(Cluster) %>% 
  summarise_each(mean) %>%
  melt(id.vars = "Cluster") %>%
  group_by(variable) %>% 
  mutate(rank = rank(value)) %>%
  ggplot(aes(x = Cluster, y = reorder(variable,rank), fill = rank)) +
  geom_tile()+
  scale_fill_distiller(palette = "Blues") +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 5)) +
  labs(y = "",fill ="Rank Ascending", title = "Figure 5: Rankings of neighborhood cluster by socio-demographic indicators in ascending order")+
  theme_bw()

```




